\chapter{Egg Implementation of the Saturation Example}\label{appendix:smce-rules}

This appendix documents the computational verification of~\autoref{example:sat-comonoid} using the \texttt{egg} equality saturation library~\cite{EggPaper}.
We describe how the symmetric monoidal theory of commutative comonoids is encoded in the \texttt{egg} framework, justify the design decisions that shape the implementation, and account for the rule application statistics in~\autoref{tab:appendix-rules}.

\paragraph{Term language.}
In \texttt{egg}, the term language is a Rust \mintinline{rust}{enum} defined via the \mintinline{rust}{define_language!} macro, which associates string tokens with constructors and fixes their arities.
\autoref{lst:language} shows the language for commutative comonoids.

\begin{listing}[h]
\caption{Term language for commutative comonoids.}
\label{lst:language}
\begin{minted}{rust}
define_language! {
    enum CommutativeComonoids {
        "mu"       = Mu,              // mu : 1 -> 2
        "i"        = I,               // i  : 1 -> 0  (counit eta)
        ";"        = Compose([Id; 2]),
        "parallel" = Parallel([Id; 2]),
        "id_0" = Identity0,  "id_1" = Identity1,
        "id_2" = Identity2,  "id_3" = Identity3,  "id_4" = Identity4,
        "sym_0_1" = Sym0_1,  "sym_1_0" = Sym1_0,  "sym_1_1" = Sym1_1,
        "sym_1_2" = Sym1_2,  "sym_2_1" = Sym2_1,  "sym_2_2" = Sym2_2,
    }
}
\end{minted}
\end{listing}

Two design decisions are visible here.

\emph{Finite enumeration of identities and symmetries.}
The SMC equations reference identities $\id_n$ and symmetries $\sym_{n,m}$ for all $n,m\in\mathbb{N}$.
Because \texttt{egg} requires a finite language, these must be enumerated at the arities actually needed by the example: identities at arities $0$--$4$ and symmetries for all $(n,m)$ with $n+m \leq 3$ and additionally $\sym_{2,2}$.
This necessitates the \emph{definitional rules} (e.g.\ $\id_2 = \id_1 \otimes \id_1$, $\sym_{2,1} = (\sym_{1,1} \otimes \id_1);(\id_1 \otimes \sym_{1,1})$) that link these tokens.
Each definitional rule is applied at most once during saturation and is omitted from~\autoref{tab:appendix-rules}.

\emph{Binary encoding of the tensor product.}
String diagrams allow $n$-ary parallel composition, but \texttt{egg} operators have fixed binary arity.
Encoding $\otimes$ as a binary \mintinline{rust}{"parallel"} operator requires explicit \emph{tensor associativity} rules $(f \otimes g) \otimes h = f \otimes (g \otimes h)$ and \emph{tensor unit} laws $\id_0 \otimes f = f$, $f \otimes \id_0 = f$ to expose all equivalent forms of a tensor expression.
As~\autoref{tab:appendix-rules} shows, these two rule groups account for more than half of all applications---a direct cost of the encoding.

\paragraph{Type analysis.}
Each morphism in an SMC has a type $(dom, cod)\in\mathbb{N}^2$.
The \texttt{egg} framework supports \emph{e-class analyses}: user-defined data associated with each equivalence class, maintained invariantly as classes are merged.
A \mintinline{rust}{TypeAnalysis} computes the type of every expression bottom-up, as shown in~\autoref{lst:types}.

\begin{listing}[h]
\caption{Type computation for composition and tensor product.}
\label{lst:types}
\begin{minted}{rust}
Compose([f, g]) => Ty { dom: get(f).dom, cod: get(g).cod },
Parallel([f, g]) => Ty {
    dom: get(f).dom + get(g).dom,
    cod: get(f).cod + get(g).cod,
},
\end{minted}
\end{listing}

Types are used to \emph{guard} rules that are only valid at specific arities.
For example, the left-identity rule $\id_n ; f = f$ must be restricted to morphisms $f$ with domain $n$:

\begin{listing}[h]
\caption{Type-guarded identity rules for $n=1$ (the pattern repeats for $n=2,3,4$).}
\label{lst:identity}
\begin{minted}{rust}
rw!("id_1-left-compose-left";  "(; id_1 ?x)" => "(?x)"
    if check_dom_type(var("?x"), 1)),
rw!("id_1-left-compose-right"; "(?x)" => "(; id_1 ?x)"
    if check_dom_type(var("?x"), 1)),
\end{minted}
\end{listing}

Without the guard, the rule would fire for morphisms of any type, producing ill-typed terms and incorrect equalities.
The guard is evaluated against the \mintinline{rust}{TypeAnalysis} data stored in each e-class.

\paragraph{Rewrite rules.}
\autoref{lst:rules} shows the remaining representative rules.
Each equation of the theory is split into two directed rules, one for each orientation, so that equality saturation can add both sides of the equivalence to the same e-class.

\begin{listing}[h]
\caption{Representative rewrite rules (simplified for presentation).}
\label{lst:rules}
\begin{minted}{rust}
// Composition associativity
rw!("assoc-compose-left";  "(; ?x (; ?y ?z))" => "(; (; ?x ?y) ?z)"),
rw!("assoc-compose-right"; "(; (; ?x ?y) ?z)" => "(; ?x (; ?y ?z))"),

// Tensor associativity
rw!("parallel-assoc-left";
    "(parallel ?x (parallel ?y ?z))" => "(parallel (parallel ?x ?y) ?z)"),
rw!("parallel-assoc-right";
    "(parallel (parallel ?x ?y) ?z)" => "(parallel ?x (parallel ?y ?z))"),

// Symmetry naturality for cod(f)=cod(g)=1
// (general form: sym_{1,1};(f*g) = (g*f);sym_{cod(g),cod(f)})
rw!("sym-nat-1-1";
    "(; sym_1_1 (parallel ?f ?g))" => "(; (parallel ?g ?f) sym_1_1)"
    if check_dom_cod(1,1, 1,1)),  // dom and cod of f and g

// Comonoid laws
rw!("comult-commutative-right"; "(; mu sym_1_1)" => "(mu)"),
rw!("comult-commutative-left";  "(mu)" => "(; mu sym_1_1)"),
rw!("counit-left"; "(; mu (parallel i id_1))" => "(id_1)"),
\end{minted}
\end{listing}

The symmetry naturality rule $\sym_{1,1};(f \otimes g) = (g \otimes f);\sym_{cod(g),cod(f)}$ cannot be expressed as a single rule because the output symmetry token depends on the codomains of $f$ and $g$, which are run-time data rather than syntactic structure.
The rule is therefore split into one instance for each possible combination of codomains, each guarded by type conditions on $f$ and $g$.
Six such instances are required for the example, covering codomains $(cod(f), cod(g)) \in \{(1,1),(1,0),(0,1),(1,2),(2,1),(2,2)\}$.

\paragraph{The verified term.}
The initial e-graph contains the term
\[
  ((\mu;\sym_{1,1});(\eta \otimes \id_1)) \otimes (\mu;(\id_1 \otimes \id_1)),
\]
represented in the egg S-expression syntax as
\begin{minted}{text}
(parallel (; (; mu sym_1_1) (parallel i id_1))
          (; mu (parallel id_1 id_1)))
\end{minted}
After saturation, the code checks that this e-class equals \mintinline{text}{(parallel id_1 mu)}, i.e.\ $\id_1 \otimes \mu$, returning \mintinline{rust}{true}.
The equivalence follows by: commutativity ($\mu;\sym_{1,1} = \mu$) followed by the counit law ($\mu;(\eta \otimes \id_1) = \id_1$) for the left component, and the right identity law ($\mu;(\id_1 \otimes \id_1) = \mu;\id_2 = \mu$) for the right component.

\paragraph{Rule application statistics.}
\autoref{tab:appendix-rules} reports the number of times each non-definitional rule was applied during saturation.

\begin{table}[h]
\centering
\begin{tabular}{@{}llp{7.5cm}rr@{}}
\toprule
Group & Rule & Equation & $(\to)$ & $(\leftarrow)$ \\
\midrule
\multirow{8}{*}{SMC}
  & Left identity       & $\id_n ; f = f$                                                                     & ---     & 899 \\
  & Right identity      & $f ; \id_n = f$                                                                     & 4       & 308 \\
  & Composition assoc.  & $(f ; g) ; h = f ; (g ; h)$                                                         & 1{,}325 & 580 \\
  & Bifunctoriality     & $(f_1 \otimes f_2) ; (g_1 \otimes g_2) = (f_1 ; g_1) \otimes (f_2 ; g_2)$           & 157     & 92  \\
  & Tensor assoc.       & $(f \otimes g) \otimes h = f \otimes (g \otimes h)$                                  & 1{,}695 & 2{,}174 \\
  & Tensor unit         & $\id_0 \otimes f = f,\quad f \otimes \id_0 = f$                                     & ---     & 2{,}005 \\
  & Sym.\ naturality    & $\sym_{1,1} ; (f \otimes g) = (g \otimes f) ; \sym$                                 & 10      & ---   \\
  & Sym.\ involution    & $\sym ; \sym = \id$                                                                  & ---     & 1     \\
\midrule
\multirow{2}{*}{$(\Sigma,\mathcal{E})$}
  & Commutativity       & $\mu ; \sym = \mu$                                                                   & 1       & --- \\
  & Counit              & $\mu ; (\eta \otimes \id) = \id$                                                     & 1       & --- \\
\midrule
  & \multicolumn{2}{l}{Total (shown rules only)} & 3{,}193 & 6{,}059 \\
\bottomrule
\end{tabular}
\caption{Rewrite rules applied during saturation in~\autoref{example:sat-comonoid}. The subscript $n$ in the identity rules ranges over $\{1,2,3,4\}$. The $(\to)$ and $(\leftarrow)$ columns give application counts for the left-to-right and right-to-left orientations of each equation; --- indicates zero applications.}
\label{tab:appendix-rules}
\end{table}

\paragraph{SMC equations.}
\emph{Left} and \emph{right identity} express that $\id_n$ is a unit for sequential composition.
The strong asymmetry (899 introductions vs.\ 0 eliminations for left identity; 308 vs.\ 4 for right identity) is a hallmark of equality saturation: the algorithm adds both sides of every equation, so it frequently introduces $\id_n ; f$ as equivalent to $f$ to enable subsequent pattern matches, but rarely needs to simplify $\id_n ; f$ away.
\emph{Composition associativity} is applied 1,905 times in total; more applications go right-to-left (flattening left-nested compositions) than left-to-right, reflecting which normal form is more useful for downstream matches.
\emph{Bifunctoriality} converts between interleaved and separated compositions.
\emph{Tensor associativity} (3,869 total) and the \emph{tensor unit} (2,005 total) together account for $56\%$ of all applications, directly reflecting the overhead of the binary \mintinline{rust}{"parallel"} encoding described above.
\emph{Symmetry naturality} fires in only one direction and only 10 times, at the specific arity combinations needed to expose the commutativity of $\mu$.
\emph{Symmetry involution} ($\sym;\sym = \id$, applied once in the introducing direction) provides the converse that allows the saturation to close the loop on symmetry-related equivalences.

\paragraph{Comonoid equations.}
\emph{Commutativity} ($\mu;\sym = \mu$) and the \emph{counit law} ($\mu;(\eta \otimes \id) = \id$) are each applied exactly once, confirming that the verification terminates as soon as the two relevant equivalences are found.
Coassociativity of $\mu$ ($\mu;(\id \otimes \mu) = \mu;(\mu \otimes \id)$) is present as a rule but not applied: the particular input term does not require it.
